---
title: "Analazing NOAAs Storm Database"
author: "Adolfo Zignago"
date: "1/29/2018"
output:
  html_document: default
  pdf_document: default
---

# Analazing NOAAs Storm DataBase: Impact in public health and economics problems
#========================================

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. 

The basic goal of this analysis is to explore the NOAA Storm Database and answer some basic questions about severe weather events. 

1. Across the United States, which types of events (as indicated in the  variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Data Processing

1. Download file
```{r Download file}
fileURL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileDest = "StormData.csv.bz2"
if(!file.exists(fileDest)) {
        download.file(fileURL, fileDest, method = "curl", mode = "wb")
}
```

2. Uncompress and load the file
```{r uncompress file}
dataFile <- bzfile(fileDest)
stormData <- read.csv(dataFile, stringsAsFactors=F)
```

3. Check and verify the data
``` {r check the data}
dim(stormData)
str(stormData)
colnames(stormData)
```

```{r verify data, results='hide'}
head(stormData)
tail(stormData)
summary(stormData)
```


### Question 1: Across the United States, which types of events (as indicated in the  variable) are most harmful with respect to population health?


### Data Analysis
```{r Subset Fatalities and Injuries}
library(ggplot2)
stormDataDS <- stormData[, c("EVTYPE","FATALITIES","INJURIES")]
stormDataPH <- aggregate(stormDataDS[, c("FATALITIES", "INJURIES")], by=list(stormDataDS$EVTYPE), FUN = sum, na.rm = T)
colnames(stormDataPH)[1] <- c("EVTYPE")

stormDataPHbyFatalities <- stormDataPH[order(-stormDataPH[, "FATALITIES"]), ][1:10, ]
stormDataPHbyInjuries <- stormDataPH[order(-stormDataPH[, "INJURIES"]), ][1:10, ]

```

- The `r stormDataPHbyFatalities[1, "EVTYPE"]` is the most event with `r stormDataPHbyFatalities[1, "FATALITIES"]` fatalities

```{r Plot Fatalities, results = 'hide'}
fatalities <- ("./images/fatalities.png")
png(fatalities, height = 480, width = 480)
ggplot(stormDataPHbyFatalities, aes(x = EVTYPE, y = FATALITIES)) + 
    geom_bar(stat = "identity", fill = "blue") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Total") + ggtitle("Top 10 Weather Events FATALITIES")
dev.off()
```

```{r echo=FALSE, out.width='50%'}
knitr::include_graphics(fatalities)
```

- The `r stormDataPHbyInjuries[1, "EVTYPE"]` is the most event with `r as.character(stormDataPHbyInjuries[1, "INJURIES"])` injuries

```{r Plot Injuries, results = 'hide'}
injuries <- ("./images/injuries.png")
png(injuries, height = 480, width = 480)
ggplot(stormDataPHbyInjuries, aes(x = EVTYPE, y = INJURIES)) + 
    geom_bar(stat = "identity", fill = "green") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Total") + ggtitle("Top 10 Weather Events INJURIES")
dev.off()
```

```{r echo=FALSE, out.width='50%'}
knitr::include_graphics(injuries)
```

``` {r Show plot1 with the analysis,  results = 'hide' }
stormDataPHTotal <- with(stormDataDS, (aggregate(FATALITIES + INJURIES ~ EVTYPE, data = stormDataDS, FUN=sum, na.rm=T)))
colnames(stormDataPHTotal) <- c("EVTYPE", "TOTAL")
stormDataPHFI <- stormDataPHTotal[order(-stormDataPHTotal[, "TOTAL"]), ][1:10, ]

plot1 <- ("./images/plot1.png")
png(plot1, height = 480, width = 480)
ggplot(stormDataPHFI, aes(x = EVTYPE, y = TOTAL)) + 
    geom_bar(stat = "identity", fill = "red") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Total") + ggtitle("Top 10 Weather Events (FATALITIES + INJURIES)")
dev.off()

```

### Results

- The `r stormDataPHbyInjuries[1, "EVTYPE"]` is the most event with `r as.character(stormDataPHFI[1, "TOTAL"])` fatalities + injuries

```{r echo=FALSE, out.width='50%'}
knitr::include_graphics(plot1)
```

### Question 2: Across the United States, which types of events have the greatest economic consequences?

The following point are the same as the question 1.
### Data Processing
1. Download file
2. Uncompress and load the file
3. Check and verify the data

### Data Analysis
```{r subset of CROPDMGEXP and PROPDMGEXP as metioned on pdf National Weather ServiceInstruction}
stormDataDS <- stormData[,c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
#Calculate the values for Property Damages
stormDataDS$PROPDMGNUM = 0
stormDataDS[stormDataDS$PROPDMGEXP == "H", ]$PROPDMGNUM = stormDataDS[stormDataDS$PROPDMGEXP == "H", ]$PROPDMG * 10^2
stormDataDS[stormDataDS$PROPDMGEXP == "K", ]$PROPDMGNUM = stormDataDS[stormDataDS$PROPDMGEXP == "K", ]$PROPDMG * 10^3
stormDataDS[stormDataDS$PROPDMGEXP == "M", ]$PROPDMGNUM = stormDataDS[stormDataDS$PROPDMGEXP == "M", ]$PROPDMG * 10^6
stormDataDS[stormDataDS$PROPDMGEXP == "B", ]$PROPDMGNUM = stormDataDS[stormDataDS$PROPDMGEXP == "B", ]$PROPDMG * 10^9


# Calculate values for Crop Damage
stormDataDS$CROPDMGNUM = 0
stormDataDS[stormDataDS$CROPDMGEXP == "H", ]$CROPDMGNUM = stormDataDS[stormDataDS$CROPDMGEXP == "H", ]$CROPDMG * 10^2
stormDataDS[stormDataDS$CROPDMGEXP == "K", ]$CROPDMGNUM = stormDataDS[stormDataDS$CROPDMGEXP == "K", ]$CROPDMG * 10^3
stormDataDS[stormDataDS$CROPDMGEXP == "M", ]$CROPDMGNUM = stormDataDS[stormDataDS$CROPDMGEXP == "M", ]$CROPDMG * 10^6
stormDataDS[stormDataDS$CROPDMGEXP == "B", ]$CROPDMGNUM = stormDataDS[stormDataDS$CROPDMGEXP == "B", ]$CROPDMG * 10^9


# Aggregate the sum of both Property and Crop Damage
stormDataETotal <- with(stormDataDS, (aggregate(PROPDMGNUM + CROPDMGNUM ~ EVTYPE, data = stormDataDS, FUN=sum, na.rm=T)))
colnames(stormDataETotal) <- c("EVTYPE", "TOTAL")

stormDataEPC <- stormDataETotal[order(-stormDataETotal[, "TOTAL"]), ][1:10, ]
```

```{r Show plot2 with the analysis, results = 'hide'}
plot2 <- ("./images/plot2.png")
png(plot2, height = 480, width = 480)
ggplot(stormDataEPC, aes(x = EVTYPE, y = (TOTAL)/10^9)) + 
    geom_bar(stat = "identity", fill = "red") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Total Damage US$ Billion") + ggtitle("Top 10 Weather Events (PROPERTY AND CROP Damage)")
dev.off()
```

### Results

- The `r stormDataEPC[1, "EVTYPE"]` is the most event with `r round(stormDataEPC[1, "TOTAL"]/10^9,2)` US$ Billion in Property and Crop damage

```{r echo=FALSE, out.width='50%'}
knitr::include_graphics(plot2)
```

